Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-13 paper.Rmd
Reduce Distance Estimation section

In the previous sections, we have defined a distance measure given in Equation \ref{eq:kl-0} for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. However, this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

We use a computer vision model to estimate this distance measure for a residual plot. Let $D$ be the result of Equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as

@@

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels in three channels (RGB), $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the estimated distance $\hat{D}$, we can compare the underlying distribution of the residuals to a theoretically "good" residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

It is not expected that $\hat{D}$ will be equal to original distance $D$. This is largely because information contained in a single residual plot is limited and it may not be able to summarize all the important characteristics of the residual distribution. For a given residual distribution $P$, many different residual plots can be simulated, where many will share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$. This suggests the error of the estimation will vary depends on whether the input residual plot is representative or not.

>>>>

We previously defined a distance measure (Equation \ref{eq:kl-0}) to quantify the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. However, this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to do a residual diagnostic in the first place.

To approximate this distance from a residual plot, we proposed a computer vision estimator $\hat{D}$ formulated as

@@

\noindent where $V_{h \times w}(.)$ renders the residuals vs fitted values plot as an $h \times w$ RGB image, and $f_{CV}(.)$ is a model that maps the image to an estimated distance $\hat{D} \in [0, +\infty)$.

The distance estimator $\hat{D}$ allows us to assess how closely the residuals resemble an ideal distribution and use as an index of model violation severity. However, $\hat{D}$ is not expected to equal the true distance $D$, as a single residual plot may not capture all characteristics of the residual distribution. For the same distribution $P$, simulated plots can vary visually, especially with small $n$, leading to estimation error that depends on how representative the input plot is.

### 2025-06-13 paper.tex
Reduce Distance Estimation section

Given the input is a single fixed-resolution image, the model output can
be defined for various tasks, including binary classification,
multiclass classification, or numeric regression.

@@

In the previous sections, we have defined a distance measure given in
Equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
However, this distance measure can only be computed when the data
generating process is known. In reality, we often have no knowledge
about the data generating process, otherwise we do not need to do a
residual diagnostic in the first place.

We use a computer vision model to estimate this distance measure for a
residual plot. Let \(D\) be the result of Equation \ref{eq:kl-0}, and
our estimator \(\hat{D}\) is formulated as

@@

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an image with
\(h \times w\) pixels in three channels (RGB), \(f_{CV}(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

With the estimated distance \(\hat{D}\), we can compare the underlying
distribution of the residuals to a theoretically ``good'' residual
distribution. \(\hat{D}\) can also be used as an index of the model
violations indicating the strength of the visual signal embedded in the
residual plot.

It is not expected that \(\hat{D}\) will be equal to original distance
\(D\). This is largely because information contained in a single
residual plot is limited and it may not be able to summarize all the
important characteristics of the residual distribution. For a given
residual distribution \(P\), many different residual plots can be
simulated, where many will share similar visual patterns, but some of
them could be visually very different from the rest, especially for
regression models with small \(n\). This suggests the error of the
estimation will vary depends on whether the input residual plot is
representative or not.

>>>>

Given the input is a single fixed-resolution image, the model output can
be defined for various tasks, including binary classification,
multiclass classification, and numeric regression.

@@

We previously defined a distance measure (Equation \ref{eq:kl-0}) to
quantify the difference between the true residual distribution \(P\) and
an ideal reference distribution \(Q\). However, this distance measure
can only be computed when the data generating process is known. In
reality, we often have no knowledge about the data generating process,
otherwise we do not need to do a residual diagnostic in the first place.

To approximate this distance from a residual plot, we proposed a
computer vision estimator \(\hat{D}\) formulated as

@@

\noindent where \(V_{h \times w}(.)\) renders the residuals vs fitted
values plot as an \(h \times w\) RGB image, and \(f_{CV}(.)\) is a model
that maps the image to an estimated distance
\(\hat{D} \in [0, +\infty)\).

The distance estimator \(\hat{D}\) allows us to assess how closely the
residuals resemble an ideal distribution and use as an index of model
violation severity. However, \(\hat{D}\) is not expected to equal the
true distance \(D\), as a single residual plot may not capture all
characteristics of the residual distribution. For the same distribution
\(P\), simulated plots can vary visually, especially with small \(n\),
leading to estimation error that depends on how representative the input
plot is.
